---
title: "Extract environmental data to GKR occurrences"
author: "Maxwell Pepperdine"
format: html
editor: visual
date: last-modified
editor_options: 
  chunk_output_type: console
---

## Overview

This script acquires giant kangaroo rat (GKR) occurrence points from GBIF using the `rgbif` package, and then extracts environmental data to the GKR occurrence points for use in MaxEnt modeling. The environmental data includes:

-   BCM climate (AET, CWD, PPT)
-   gSSURGO soil data (pH, % clay, % silt, EC)
-   Slope from USGS Open Topography

We will also spatially thin the occurrence points by 1 km to reduce spatial autocorrelation.

```{r}
rm(list=ls())
```

## Load packages

```{r}
library(terra)      # for raster processing
library(tidyverse)  # always
library(sf)         # for spatial data processing
library(here)       # for file paths
library(spThin)     # for spatial thinning
library(rgbif)      # for GBIF data download
```

## Load data

#### Environmental rasters

```{r}
# load all env rasters
env_rasters <- rast(list.files(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline"), 
                               pattern = ".tif$", 
                               full.names = TRUE))
# fix names
names(env_rasters) <- c("aet1991_2020_ave", "clay_pct_0_10cm", "cwd1991_2020_ave", 
                        "ec_dS_m_0_10cm", "pH_0_10cm", "ppt1991_2020_ave", 
                        "silt_pct_0_10cm", "slope")
plot(env_rasters)
```

#### GKR occurrences

```{r}
# load occurrence points from GBIF
gkr_occs <- occ_search(
  taxonKey = 2439550,  # taxon key for Dipodomys ingens
  hasCoordinate = TRUE,
  hasGeospatialIssue = FALSE,
  year = "1990,2025", 
  basisOfRecord = "HUMAN_OBSERVATION;OBSERVATION;MACHINE_OBSERVATION",
  limit = 10000
)

# convert to data frame
gkr_occ_df <- gkr_occs$data %>%
  dplyr::select(gbifID, decimalLongitude, decimalLatitude, institutionCode, year, 
         basisOfRecord, scientificName)

# convert occ points to SpatVector
gkr_occ_vect <- vect(gkr_occ_df, 
                     geom = c("decimalLongitude", "decimalLatitude"), 
                     crs = "EPSG:4326")

# make sure crs matches env rasters
gkr_occ_vect <- project(gkr_occ_vect, crs(env_rasters))
```

#### Spatial thinning of occurrence points

Thin occurrence points to 1 km minimum distance. This reduces spatial autocorrelation and sampling bias.

```{r}
# convert back to lat/lon for spThin (it requires decimal degrees)
gkr_occ_latlon <- project(gkr_occ_vect, "EPSG:4326")
gkr_occ_df <- as.data.frame(gkr_occ_latlon, geom = "XY")

# run spatial thinning with 1 km distance
# thin.par = 1 means 1 km minimum distance between points
# reps = 100 means run 100 iterations to find best thinning solution
# locs.thinned.list.return = TRUE returns all replicates
# write.files = FALSE prevents writing files automatically
thinned_data <- spThin::thin(
  loc.data = gkr_occ_df,
  lat.col = "y",  # latitude column
  long.col = "x", # longitude column
  spec.col = "scientificName", 
  thin.par = 1,   # thinning distance in km
  reps = 100,     # number of replications
  locs.thinned.list.return = TRUE,
  write.files = FALSE,
  write.log.file = FALSE
)

# select the thinned dataset with the most points
# (spThin returns a list of different thinning solutions)
max_points <- which.max(sapply(thinned_data, nrow))
gkr_thinned_df <- thinned_data[[max_points]]

# convert thinned points back to SpatVector and match original CRS
gkr_thinned_vect <- vect(gkr_thinned_df, 
                          geom = c("Longitude", "Latitude"), 
                          crs = "EPSG:4326")

# project to match env rasters
gkr_thinned_vect <- project(gkr_thinned_vect, crs(env_rasters))
crs(gkr_thinned_vect)  # check crs
```

## Extract environmental data to occurrence points

```{r}
# extract env data to occ points
gkr_occ_env <- terra::extract(env_rasters, gkr_thinned_vect, 
                               bind = TRUE)

# set as sf object
gkr_occ_env_sf <- st_as_sf(gkr_occ_env, 
                            coords = c("longitude", "latitude"))

# extract geometry columns as lat/lon
gkr_occ_env_df <- gkr_occ_env_sf %>% 
  mutate(longitude = st_coordinates(gkr_occ_env_sf)[,1],
         latitude = st_coordinates(gkr_occ_env_sf)[,2], 
         .before = "aet1991_2020_ave") %>%
  st_drop_geometry()
```

## Save the data

```{r}
# save final occurrence points with env data to CSV
write_csv(gkr_occ_env_df, here("data/intermediate/misc/habitat_suitability/sdm_files/gkr_sdm/1_extract_occ_env_data_gkr/gkr_occ_env_data.csv"))

# save SF object
write_sf(gkr_occ_env_sf, here("data/intermediate/misc/habitat_suitability/sdm_files/gkr_sdm/1_extract_occ_env_data_gkr/gkr_occ_env_data.shp"),
         append = FALSE)
```
