---
title: "Extract environmental data to GKR occurrences"
format: html
editor: visual
date: last-modified
editor_options: 
  chunk_output_type: console
---

## Overview

This script acquires giant kangaroo rat (GKR) occurrence points from GBIF using the `rgbif` package, and then extracts environmental data to the GKR occurrence points for use in MaxEnt modeling. The environmental data includes:

-   BCM climate (AET, CWD, PPT, TMN, TMX)
-   gSSURGO soil data (pH, % clay, % silt, % sand, EC)
-   Slope from USGS Open Topography

We will also spatially thin the occurrence points to one per 270 m raster cell to reduce spatial autocorrelation and match the resolution of the environmental predictors.

```{r}
rm(list=ls())
```

## Load packages

```{r}
library(terra)      # for raster processing
library(tidyverse)  # always
library(sf)         # for spatial data processing
library(here)       # for file paths
library(rgbif)      # for GBIF data download
```

## Load data

#### Environmental rasters

```{r}
# load all env rasters
env_rasters <- rast(list.files(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline"), 
                               pattern = ".tif$", 
                               full.names = TRUE))
# fix names
names(env_rasters) <- c("aet1991_2020_ave", "clay_pct_0_10cm", "cwd1991_2020_ave", 
                        "ec_dS_m_0_10cm", "pet1991_2020_ave", "pH_0_10cm", "ppt1991_2020_ave", 
                        "sand_pct_0_10cm", "silt_pct_0_10cm", "slope", 
                        "tmn1991_2020_ave", "tmx1991_2020_ave")
# plot(env_rasters)
```

#### GKR occurrences

```{r}
# load occurrence points from GBIF
gkr_occs <- occ_search(
  taxonKey = 2439550,  # taxon key for Dipodomys ingens
  hasCoordinate = TRUE,
  hasGeospatialIssue = FALSE,
  year = "1990,2025", 
  basisOfRecord = "HUMAN_OBSERVATION;OBSERVATION",
  limit = 10000
)

# convert to data frame
gkr_occ_df <- gkr_occs$data %>%
  dplyr::select(gbifID, decimalLongitude, decimalLatitude, institutionCode, year, 
         basisOfRecord, scientificName)

# convert occ points to SpatVector
gkr_occ_vect <- vect(gkr_occ_df, 
                     geom = c("decimalLongitude", "decimalLatitude"), 
                     crs = "EPSG:4326")

# make sure crs matches env rasters
gkr_occ_vect <- project(gkr_occ_vect, crs(env_rasters))
```

#### Spatial thinning of occurrence points

Thin occurrence points to one per 270 m grid cell. This reduces spatial autocorrelation and sampling bias, and matches the resolution of our environmental data.

```{r}
# Thin occurrence points to one per raster cell (270 m)
# This ensures thinning matches the resolution of the environmental predictors

# get the raster cell ID for each occurrence point
cell_ids <- terra::cellFromXY(env_rasters[[1]], crds(gkr_occ_vect))

# shuffle points so the "kept" point per cell is random
set.seed(123)
random_order <- sample(seq_len(nrow(gkr_occ_vect)))
gkr_occ_shuffled <- gkr_occ_vect[random_order, ]
cell_ids_shuffled <- cell_ids[random_order]

# keep first (now random) point per cell
gkr_thinned_vect <- gkr_occ_shuffled[!duplicated(cell_ids_shuffled), ]
crs(gkr_thinned_vect)
```

## Extract environmental data to occurrence points

```{r}
# extract env data to occ points
gkr_occ_env <- terra::extract(env_rasters, gkr_thinned_vect, 
                               bind = TRUE)

# set as sf object
gkr_occ_env_sf <- st_as_sf(gkr_occ_env, 
                            coords = c("longitude", "latitude"))

# extract geometry columns as lat/lon
gkr_occ_env_df <- gkr_occ_env_sf %>% 
  mutate(longitude = st_coordinates(gkr_occ_env_sf)[,1],
         latitude = st_coordinates(gkr_occ_env_sf)[,2], 
         .before = "aet1991_2020_ave") %>%
  st_drop_geometry() %>% 
  dplyr::select(6:19)
```

## Save the data

```{r}
# save final occurrence points with env data to CSV
write_csv(gkr_occ_env_df, here("data/intermediate/misc/habitat_suitability/sdm_files/gkr_sdm/1_extract_occ_env_data_gkr/gkr_occ_env_data.csv"))
```
