---
title: "Generate GKR background points"
author: "Maxwell Pepperdine"
format: html
editor: visual
date: last-modified
editor_options: 
  chunk_output_type: console
---

```{r}
# clean environment
rm(list = ls())
```

## Overview

This script generates the bg data for habitat suitability modeling for the giant kangaroo rat (*Dipodomys ingens*) in the San Joaquin Valley, CA. It will load our GKR occurrence data, generate a 50 km buffer around each occurrence point, and then find all geographically unique vertebrate (phylum Vertebrata) occurrence points within that buffer. These will be the background points for the GKR MaxEnt model.

## Load packages

```{r}
library(tidyverse)  # always
library(here)       # for file paths
library(rgbif)      # for GBIF data download
library(sf)         # for spatial data processing
```

## Load data

```{r}
# Load GKR occurrence data
gkr_occs <- read_csv(here("data/intermediate/misc/habitat_suitability/sdm_files/gkr_sdm/1_extract_occ_env_data_gkr/gkr_occ_env_data.csv"))
```

## Data processing

#### Create 50 km buffer around GKR occ points

```{r}
# convert GKR occ data to sf object
gkr_sf <- st_as_sf(gkr_occs, 
                    coords = c("longitude", "latitude"), 
                    crs = 3310)

# create 50 km buffer around each occurrence point
gkr_buffer <- st_buffer(gkr_sf, dist = 50000)

# combine buffers into a single polygon
gkr_buffer_union <- st_union(gkr_buffer)

# transform buffer back to WGS84
gkr_buffer_wgs84 <- st_transform(gkr_buffer_union, crs = 4326)
```

```{r}
# set buffer to WKT format for rgbif::occ_download()
gkr_buffer_wkt <- st_as_text(st_geometry(gkr_buffer_wgs84)[[1]])
nchar(gkr_buffer_wkt)  # check length of WKT string

# buffer is too long for rgbif, so simplify it
buffer_simplify <- st_simplify(gkr_buffer_wgs84, dTolerance = 500)
buffer_simplify_wkt <- st_as_text(st_geometry(buffer_simplify)[[1]])
nchar(buffer_simplify_wkt)  # check length of WKT string
```

#### Obtain & process occ records using `rgbif::occ_download()`

```{r}
# set up GBIF credentials (one-time setup)
options(gbif_user = "mpepperdine")
options(gbif_pwd = " ")
options(gbif_email = " ")

# use rgbif::occ_download to acquire background points
download_key <- occ_download(
  pred_in("taxonKey", 44),  # Vertebrata
  pred("hasCoordinate", TRUE),
  pred("hasGeospatialIssue", FALSE),
  pred_gte("year", 1990),
  pred_lte("year", 2025),
  pred_lte("coordinateUncertaintyInMeters", 100),
  pred_in("basisOfRecord", c("HUMAN_OBSERVATION", "MACHINE_OBSERVATION", "OBSERVATION")),
  pred_within(buffer_simplify_wkt)  # get bg points within the buffer
)

# check download status
cat("Download key:", download_key, "\n")
cat("Check status at: https://www.gbif.org/occurrence/download/", download_key, "\n")

# check the download status programatically; this might take a while...
occ_download_wait(download_key, status_ping = 10)  # Check every 10 seconds

# download the data
download_file <- occ_download_get(download_key, path = ".")

# import the data
vertebrate_data <- occ_download_import(download_file)

# convert to sf
vertebrate_sf <- st_as_sf(vertebrate_data,
                          coords = c("decimalLongitude", "decimalLatitude"),
                          crs = 4326) %>% 
  # remove all GKR occ points
  filter(species != "Dipodomys ingens") %>% 
  filter(taxonKey != 2439550)


# intersect occ points with the original precise buffer
# this keeps only points within 50 km of GKR occs
vertebrate_in_buffer <- st_intersection(vertebrate_sf, gkr_buffer_wgs84)

# remove duplicates
vertebrate_coords <- vertebrate_in_buffer %>%
  st_drop_geometry() %>%
  bind_cols(
    st_coordinates(vertebrate_in_buffer) %>%
      as.data.frame() %>%
      rename(longitude = X, latitude = Y)
  )

# de-duplicate on numeric coords
background_df <- vertebrate_coords %>%
  distinct(longitude, latitude, .keep_all = TRUE) %>%
  dplyr::select(gbifID, institutionCode, year, basisOfRecord, 
                scientificName, class, classKey, species,
                longitude, latitude)
```

Save the data

```{r}
# save final background points to CSV
write_csv(background_df, here("data/intermediate/misc/habitat_suitability/sdm_files/gkr_sdm/2_generate_bg_points_gkr/gkr_bg_points.csv"))
```
