---
title: "Extract environmental data to GKR background points"
format: html
editor: visual
date: last-modified
editor_options: 
  chunk_output_type: console
---

## Overview

This script extracts environmental data to the GKR background points (created in `2_generate_bg_points_gkr.qmd`) for use in MaxEnt modeling. The environmental data includes:

-   BCM climate (AET, CWD, PPT)
-   gSSURGO soil data (pH, % clay, % silt, EC)
-   Slope from USGS Open Topography

```{r}
rm(list=ls())
```

## Load packages

```{r}
library(terra)      # for raster processing
library(tidyverse)  # always
library(sf)         # for spatial data processing
library(here)       # for file paths
```

## Load data

#### Environmental rasters

```{r}
# load all env rasters
env_rasters <- rast(list.files(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline"), 
                               pattern = ".tif$", 
                               full.names = TRUE))
# fix names
names(env_rasters) <- c("aet1991_2020_ave", "clay_pct_0_10cm", "cwd1991_2020_ave", 
                        "ec_dS_m_0_10cm", "pH_0_10cm", "ppt1991_2020_ave", 
                        "silt_pct_0_10cm", "slope")
plot(env_rasters)
```

#### GKR background points (created in `2_generate_bg_points_gkr.qmd`)

```{r}
# load background points
gkr_bg <- read_csv(here("data/intermediate/misc/habitat_suitability/sdm_files/gkr_sdm/2_generate_bg_points_gkr/gkr_bg_points.csv"))

# convert bg points to SpatVector
gkr_bg_vect <- vect(gkr_bg, 
                     geom = c("longitude", "latitude"), 
                     crs = "EPSG:4326")

# make sure crs matches env rasters
gkr_bg_vect <- project(gkr_bg_vect, crs(env_rasters))
crs(gkr_bg_vect)
```

#### Spatial thinning of background points (1 km)

Thin background points by 1 km to reduce spatial autocorrelation.

```{r}
# # convert back to lat/lon for spThin (it requires decimal degrees)
# gkr_bg_latlon <- project(gkr_bg_vect, "EPSG:4326")
# gkr_bg_df <- as.data.frame(gkr_bg_latlon, geom = "XY")
# # add species dummy column for spThin
# gkr_bg_df$species <- "background"
# 
# # run spatial thinning with 1 km distance
# thinned_data <- spThin::thin(
#   loc.data = gkr_bg_df,
#   lat.col = "y",  # latitude column
#   long.col = "x", # longitude column
#   spec.col = "species", # dummy species column
#   thin.par = 1,   # thinning distance in km
#   reps = 10,     # number of replications
#   locs.thinned.list.return = TRUE,
#   write.files = FALSE,
#   write.log.file = FALSE
# )
# 
# # select the thinned dataset with the most points
# max_points <- which.max(sapply(thinned_data, nrow))
# gkr_bg_thinned_df <- thinned_data[[max_points]]
# 
# # convert thinned points back to SpatVector and match original CRS
# gkr_bg_thinned_vect <- vect(gkr_bg_thinned_df, 
#                              geom = c("Longitude", "Latitude"), 
#                              crs = "EPSG:4326")
# 
# # project to match env rasters
# gkr_bg_thinned_vect <- project(gkr_bg_thinned_vect, crs(env_rasters))
# crs(gkr_bg_thinned_vect)  # check crs
```

Having RAM issues when using `spThin` on larger datasets, so using grid-based thinning instead.

```{r}
# convert back to lat/lon
gkr_bg_latlon <- project(gkr_bg_vect, "EPSG:4326")
gkr_bg_df <- as.data.frame(gkr_bg_latlon, geom = "XY")

# convert to sf object
gkr_bg_sf <- st_as_sf(gkr_bg_df, 
                      coords = c("x", "y"), 
                      crs = 4326)

# project to EPSG 3310 for distance calculations
gkr_bg_projected <- st_transform(gkr_bg_sf, crs = 3310)

# create a grid ID based on 1km cells
grid_size <- 1000  # 1 km in meters

gkr_bg_projected <- gkr_bg_projected %>%
  mutate(
    grid_x = floor(st_coordinates(.)[,1] / grid_size),
    grid_y = floor(st_coordinates(.)[,2] / grid_size),
    grid_id = paste(grid_x, grid_y, sep = "_")
  )

# keep one random point per grid cell
set.seed(123)  # for reproducibility
gkr_bg_thinned_sf <- gkr_bg_projected %>%
  group_by(grid_id) %>%
  slice_sample(n = 1) %>%
  ungroup() %>%
  dplyr::select(-grid_x, -grid_y, -grid_id)

# transform back to WGS84
gkr_bg_thinned_sf <- st_transform(gkr_bg_thinned_sf, crs = 4326)

# convert thinned points back to SpatVector and match original CRS
gkr_bg_thinned_vect <- vect(gkr_bg_thinned_sf)

# project to match env rasters
gkr_bg_thinned_vect <- project(gkr_bg_thinned_vect, crs(env_rasters))
crs(gkr_bg_thinned_vect)  # check crs
```

## Extract environmental data to background points

```{r}
# extract env data to bg points
gkr_bg_env <- terra::extract(env_rasters, gkr_bg_thinned_vect, 
                              bind = TRUE)

# set as sf object
gkr_bg_env_sf <- st_as_sf(gkr_bg_env, 
                           coords = c("longitude", "latitude")) %>% 
  dplyr::select(all_of(names(env_rasters)))

# extract geometry columns as lon/lat
gkr_bg_env_df <- gkr_bg_env_sf %>% 
  mutate(longitude = st_coordinates(gkr_bg_env_sf)[,1],
         latitude = st_coordinates(gkr_bg_env_sf)[,2], 
         .before = 1) %>% 
  st_drop_geometry()
```

## Save the data

```{r}
# save final background points with env data to CSV
write_csv(gkr_bg_env_df, here("data/intermediate/misc/habitat_suitability/sdm_files/gkr_sdm/3_extract_bg_env_data_gkr/gkr_bg_env_data.csv"))

# save SF object
write_sf(gkr_bg_env_sf, here("data/intermediate/misc/habitat_suitability/sdm_files/gkr_sdm/3_extract_bg_env_data_gkr/gkr_bg_env_data.shp"),
         append = FALSE)
```
