---
title: "GKR habitat suitability prediction plotting"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
# clear the environment
rm(list = ls())
```

## Overview/Purpose

This script generates maps of the GKR habitat suitability predictions.

## Setup

#### Load Packages

```{r}
library(terra)
library(tidyverse)
library(here)
library(sf)
```

#### Load data

###### Habitat suitability

```{r}
##### giant kangaroo rat habitat suitability rasters #####


```

###### Misc

```{r}
# load SJV boundary for masking
sjv_boundary <- read_sf(here("data/raw/sjv_counties/sjv_counties.shp"))
```

## Alignments checks

Raster math requires matching CRS, extent, and resolution.

```{r}

```

#### Make sure SJV is in `EPSG:3310`

```{r}
# transform SJV boundary to EPSG:3310
sjv_boundary <- st_transform(sjv_boundary, crs = crs(bnll_fallowSJV))
crs(sjv_boundary)
# convert to terra vector
sjv_vect <- vect(sjv_boundary)
```

## Plotting

#### Continuous (raw data) comparisons

Side-by-side plot of the two raw models

```{r}

```

Difference raster map. 

```{r}

```

#### Reclassified comparisons

Threshold: 

```{r}

```

Side-by-side plot of the two binary models

```{r}

```

Difference raster map (binary).

```{r}

```








