---
title: "Getting started with prioritizR"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

This tutorial provides a short introduction to the `prioritizr` R package, sourced from <https://prioritizr.net/articles/prioritizr.html#introduction> 

## Load packages

```{r}
library(prioritizrdata)
library(prioritizr)
library(sf)
library(terra)
library(vegan)
library(cluster)
```

## Load data

```{r}
# set seed for reproducibility
set.seed(500)

# load planning unit data
tas_pu <- prioritizrdata::get_tas_pu()

# load feature data
tas_features <- prioritizrdata::get_tas_features()
```

## Data exploration and processing

```{r}
# print planning unit data
print(tas_pu)

# set costs for existing protected areas to 0
tas_pu$cost <- tas_pu$cost * !tas_pu$locked_in

# plot map of planning unit costs
plot(st_as_sf(tas_pu[, "cost"]), main = "Planning unit costs")

# plot map of planning unit coverage by protected areas 
plot(st_as_sf(tas_pu[, "locked_in"]), main = "Protected area coverage")
```

```{r}
# print the conservation feature data 
print(tas_features)

# plot a map of the first four vegetation types
plot(tas_features[[1:4]])
```

## Problem formulation

```{r}
# build problem
p1 <-
  problem(tas_pu, tas_features, cost_column = "cost") %>%
  add_min_set_objective() %>%
  add_boundary_penalties(penalty = 0.005) %>%
  add_relative_targets(0.17) %>%
  add_locked_in_constraints("locked_in") %>%
  add_binary_decisions() %>% 
  add_gurobi_solver()

# print problem
print(p1)
```

## Prioritization

```{r}
# solve problem
s1 <- solve(p1)
```

```{r}
# plot map of prioritization
plot(
  st_as_sf(s1[, "solution_1"]), main = "Prioritization",
  pal = c("grey90", "darkgreen")
)
```

## Feature representation

Let’s examine how well the vegetation communities are represented by existing protected areas and the prioritization

```{r}
# create column with existing protected areas
tas_pu$pa <- round(tas_pu$locked_in)

# calculate feature representation statistics based on existing protected areas
tc_pa <- eval_target_coverage_summary(p1, tas_pu[, "pa"])
print(tc_pa)
```

```{r}
# calculate  feature representation statistics based on the prioritization
tc_s1 <- eval_target_coverage_summary(p1, s1[, "solution_1"])
print(tc_s1)
```

```{r}
# explore representation by existing protected areas
## calculate number of features adequately represented by existing protected
## areas
sum(tc_pa$met)
```

```{r}
## summarize representation (values show percent coverage)
summary(tc_pa$relative_held * 100)
```

```{r}
## visualize representation  (values show percent coverage)
hist(tc_pa$relative_held * 100,
     main = "Feature representation by existing protected areas",
     xlim = c(0, 100),
     xlab = "Percent coverage of features (%)")
```

```{r}
# explore representation by prioritization
## summarize representation (values show percent coverage)
summary(tc_s1$relative_held * 100)
```

```{r}
## calculate number of features adequately represented by the prioritization
sum(tc_s1$met)
```

```{r}
## visualize representation  (values show percent coverage)
hist(
  tc_s1$relative_held * 100,
  main = "Feature representation by prioritization",
  xlim = c(0, 100),
  xlab = "Percent coverage of features (%)"
)
```

## Evaluating importance

```{r}
# calculate relative importance
imp_s1 <- eval_rank_importance(p1, s1["solution_1"], n = 10)
print(imp_s1)

# manually set locked in planning units to -1 to help with visualization,
# this way we can easily see the importance scores for the priority areas
imp_s1$rs[tas_pu$locked_in] <- -1
```

```{r}
# plot map of importance scores
plot(st_as_sf(imp_s1[, "rs"]), main = "Overall importance")
```

## Portfolios

```{r}

```
















