---
title: "Processing Env Predictor Variables for SDMs"
format: html
---

```{r}
# clear the environment
rm(list = ls())
```

## Overview/Purpose

This script processes environmental predictor variables for use in species distribution models (SDMs). It will load each of the predictor variables in terra's `SpatRaster` format, ensure they are aligned to a common extent and resolution, and then export the processed rasters for use in modeling.

## Setup

#### Load packages

```{r}
# load required packages
library(terra)
library(here)
library(tidyverse)
library(tigris)
library(sf)
```

#### Load data

###### BCM

```{r}
# ==============================================================================
# BCM historic data (30-year averages from 1991-2020)
# ==============================================================================

## AET
aet_rast <- rast(here("data/raw/bcm/BCMv8_1991_2020_aves/aet1991_2020_ave.asc"))
crs(aet_rast) <- "EPSG:3310" # assign correct CRS
# QC resolution & crs
res(aet_rast) # 270m resolution
crs(aet_rast) # EPSG:3310

## MAP
ppt_rast <- rast(here("data/raw/bcm/BCMv8_1991_2020_aves/ppt1991_2020_ave.asc"))
crs(ppt_rast) <- "EPSG:3310" # assign correct CRS
# QC resolution & crs
res(ppt_rast) # 270m resolution
crs(ppt_rast) # EPSG:3310

## CWD
cwd_rast <- rast(here("data/raw/bcm/BCMv8_1991_2020_aves/cwd1991_2020_ave.asc"))
crs(cwd_rast) <- "EPSG:3310" # assign correct CRS
# QC resolution & crs
res(cwd_rast) # 270m resolution
crs(cwd_rast) # EPSG:3310
```

###### gSSURGO

```{r}
# ==============================================================================
# gSSURGO
# ==============================================================================

## % clay
# file path for gSSURGO % clay raster
clay_path <- here("data/raw/gssurgo/processed/clay_pct_0_10cm_CA_270m.tif")
clay_rast <- rast(clay_path)

## pH
# file path for gSSURGO pH raster
ph_path <- here("data/raw/gssurgo/processed/pH_0_10cm_CA_270m.tif")
ph_rast <- rast(ph_path)

# EC
# file path for gSSURGO EC raster
ec_path <- here("data/raw/gssurgo/processed/ec_dS_m_0_10cm_CA_270m.tif")
ec_rast <- rast(ec_path)
```

###### Slope

```{r}
# ==============================================================================
# slope (open topography)
# ==============================================================================

# open topography limits the size of raster downloads
# we have separate elevation rasters for northern and southern CA

# northern CA elevation raster path
nCA_elevation_path <- here("data/raw/open_topography/output_nCA_USGS30m.tif")
nCA_elevation_rast <- rast(nCA_elevation_path)

# southern CA elevation raster path
sCA_elevation_path <- here("data/raw/open_topography/output_sCA_USGS30m.tif")
sCA_elevation_rast <- rast(sCA_elevation_path)
```

## Data Processing

#### Make SJV spat vector

```{r}
# # get California counties from Census TIGER/Line shapefiles
# ca_counties <- counties(state = "CA")
# 
# # define San Joaquin Valley counties
# sjv_counties <- c("San Joaquin", "Stanislaus", "Merced", "Madera", 
#                   "Fresno", "Kings", "Tulare", "Kern")
# 
# # filter and combine into one polygon
# sjv <- ca_counties %>%
#   filter(NAME %in% sjv_counties) %>%
#   st_union() %>% 
#   st_transform(crs = crs(aet_mean)) %>% # transform to aet raster CRS
#   vect() # convert to terra vector format
# 
# # crop AET mean raster to SJV extent
# aet_mean_sjv <- aet_mean %>% 
#   crop(sjv) %>%
#   mask(sjv)
# 
# # plot cropped AET raster
# plot(aet_mean_sjv, main = "Mean AET (SJV)")
```

#### Make CA boundary spat vector

```{r}
# get California state boundary from Census TIGER/Line shapefiles
ca_state <- states() %>%
  filter(NAME == "California") %>%
  st_transform(crs = crs(aet_rast)) %>% # transform to aet raster CRS
  vect() # convert to terra vector format
plot(ca_state, main = "California State Boundary")

# # transform to WGS84 for later use
# ca_state_wgs84 <- states() %>%
#   filter(NAME == "California") %>%
#   st_transform(crs = "EPSG:4326") %>% # transform to WGS84
#   vect() # convert to terra vector format
# plot(ca_state_wgs84, main = "California State Boundary (WGS84)")
# 
# # save CA boundary as WGS84
# writeVector(ca_state_wgs84, 
#             filename = here("data/raw/ca_state/ca_state_boundary_wgs84.shp"), 
#             overwrite = TRUE)
```

#### BCM

###### Crop `AET`, `CWD`, and `PPT` CA boundary

```{r}
# crop AET raster to CA boundary
aet_ca <- aet_rast %>%
  crop(ca_state) %>%
  mask(ca_state)

# view range of AET values in CA
summary(values(aet_ca))
# -9999 indicates no data values -- these should be NA
values(aet_ca)[values(aet_ca) == -9999] <- NA


# crop PPT raster to CA boundary
ppt_ca <- ppt_rast %>%
  crop(ca_state) %>%
  mask(ca_state)
# view range of PPT values in CA
summary(values(ppt_ca))
# -9999 indicates no data values -- these should be NA
values(ppt_ca)[values(ppt_ca) == -9999] <- NA


# crop CWD raster to CA boundary
cwd_ca <- cwd_rast %>%
  crop(ca_state) %>%
  mask(ca_state)
# view range of CWD values in CA
summary(values(cwd_ca))
# -9999 indicates no data values -- these should be NA
values(cwd_ca)[values(cwd_ca) == -9999] <- NA


# plot cropped rasters
plot(aet_ca, main = "AET (CA Boundary)")
plot(ppt_ca, main = "PPT (CA Boundary)")
plot(cwd_ca, main = "CWD (CA Boundary)")
```

###### Resample BCM rasters to common extent and resolution

```{r}
# resample PPT to match AET extent
ppt_ca <- resample(ppt_ca, 
                   aet_ca, 
                   method = "mean")

# resample CWD to match AET extent
cwd_ca <- resample(cwd_ca, 
                   aet_ca, 
                   method = "mean")

# confirm all rasters have the same extent and resolution
crs(aet_ca) == crs(ppt_ca)
ext(aet_ca) == ext(ppt_ca)
res(aet_ca) == res(ppt_ca)

crs(aet_ca) == crs(cwd_ca)
ext(aet_ca) == ext(cwd_ca)
res(aet_ca) == res(cwd_ca)
```

#### gSSURGO

Resample gSSURGO rasters to match AET `ext`, `res`, and `crs`.

```{r}
# resample gSSURGO rasters to match AET
clay_resampled <- resample(clay_rast, 
                           aet_ca, 
                           method = "mean")
ph_resampled <- resample(ph_rast, 
                         aet_ca, 
                         method = "mean")
ec_resampled <- resample(ec_rast, 
                         aet_ca, 
                         method = "mean")

# confirm all rasters have the same extent and resolution
crs(clay_resampled) == crs(aet_ca)
crs(ph_resampled) == crs(aet_ca)
crs(ec_resampled) == crs(aet_ca)

ext(clay_resampled) == ext(aet_ca)
ext(ph_resampled) == ext(aet_ca)
ext(ec_resampled) == ext(aet_ca)

res(clay_resampled) == res(aet_ca)
res(ph_resampled) == res(aet_ca)
res(ec_resampled) == res(aet_ca)
```

#### Slope

```{r}
# project elevation rasters to match AET
nCA_elevation_proj <- project(nCA_elevation_rast, 
                              aet_ca, 
                              method = "mean")
crs(nCA_elevation_proj) # confirm EPSG:3310

sCA_elevation_proj <- project(sCA_elevation_rast, 
                              aet_ca, 
                              method = "mean")
crs(sCA_elevation_proj) # confirm EPSG:3310  

# resample southern CA elevation to match northern CA elevation
sCA_elevation_resampled <- resample(sCA_elevation_proj, 
                                    nCA_elevation_proj, 
                                    method = "mean")


# merge northern and southern CA elevation rasters
elevation_merged <- terra::merge(nCA_elevation_proj, sCA_elevation_resampled)

# crop to CA
elevation_ca <- elevation_merged %>%
  crop(ca_state) %>%
  mask(ca_state)
plot(elevation_ca, main = "Elevation (CA Boundary)")


# calculate slope from elevation
slope_rast <- terrain(elevation_ca, 
                      v = "slope", 
                      unit = "degrees")

# resample to match AET resolution and extent
slope_resampled <- resample(slope_rast, 
                            aet_ca, 
                            method = "mean")
plot(slope_resampled, main = "Elevation")


# confirm all rasters have the same extent and resolution
crs(slope_resampled) == crs(aet_ca)
ext(slope_resampled) == ext(aet_ca)
res(slope_resampled) == res(aet_ca)
```

## Export Processed Rasters

#### EPSG:3310, 270m resolution, SJV extent

```{r}
# path to output directory
output_dir <- here("data/intermediate/misc/habitat_suitability/env_predictors/baseline/keep_NAs")

# export AET raster
writeRaster(aet_ca, 
            filename = file.path(output_dir, "aet1991_2020_ave_CA_270m.tif"), 
            overwrite = TRUE)

# export PPT raster
writeRaster(ppt_ca, 
            filename = file.path(output_dir, "ppt1991_2020_ave_CA_270m.tif"), 
            overwrite = TRUE)

# export CWD raster
writeRaster(cwd_ca,
            filename = file.path(output_dir, "cwd1991_2020_ave_CA_270m.tif"), 
            overwrite = TRUE)


# export gSSURGO rasters
writeRaster(clay_resampled, 
            filename = file.path(output_dir, "clay_pct_0_10cm_CA_270m.tif"), 
            overwrite = TRUE)
writeRaster(ph_resampled, 
            filename = file.path(output_dir, "ph_0_10cm_CA_270m.tif"), 
            overwrite = TRUE)
writeRaster(ec_resampled, 
            filename = file.path(output_dir, "ec_dS_m_0_10cm_CA_270m.tif"), 
            overwrite = TRUE)


# export slope raster
writeRaster(slope_resampled, 
            filename = file.path(output_dir, "slope_deg_CA_270m.tif"), 
            overwrite = TRUE)
```

#### WGS84, ~0.00242 degree resolution at equator, SJV extent

Reproject all rasters to WGS 84 (`EPSG:4326`) for use in Wallace.

```{r}
# project all rasters to WGS 84
aet_wgs84 <- project(aet_ca, "EPSG:4326", method = "mean")
ppt_wgs84 <- project(ppt_ca, "EPSG:4326", method = "mean")
cwd_wgs84 <- project(cwd_ca, "EPSG:4326", method = "mean")
clay_wgs84 <- project(clay_resampled, "EPSG:4326", method = "mean")
ph_wgs84 <- project(ph_resampled, "EPSG:4326", method = "mean")
ec_wgs84 <- project(ec_resampled, "EPSG:4326", method = "mean")
slope_wgs84 <- project(slope_resampled, "EPSG:4326", method = "mean")

# confirm all rasters have the same extent and resolution
crs(aet_wgs84) == crs(ppt_wgs84)
crs(aet_wgs84) == crs(clay_wgs84)
crs(aet_wgs84) == crs(ph_wgs84)
crs(aet_wgs84) == crs(ec_wgs84)
crs(aet_wgs84) == crs(slope_wgs84)
crs(aet_wgs84) == crs(cwd_wgs84)

ext(aet_wgs84) == ext(ppt_wgs84)
ext(aet_wgs84) == ext(clay_wgs84)
ext(aet_wgs84) == ext(ph_wgs84)
ext(aet_wgs84) == ext(ec_wgs84)
ext(aet_wgs84) == ext(slope_wgs84)
ext(aet_wgs84) == ext(cwd_wgs84)

res(aet_wgs84) == res(ppt_wgs84)
res(aet_wgs84) == res(clay_wgs84)
res(aet_wgs84) == res(ph_wgs84)
res(aet_wgs84) == res(ec_wgs84)
res(aet_wgs84) == res(slope_wgs84)
res(aet_wgs84) == res(cwd_wgs84)
```

Save the WGS84 rasters.

```{r}
# export BCM rasters in WGS84
writeRaster(aet_wgs84, 
            filename = file.path(output_dir, "WGS84/aet1991_2020_ave_CA_wgs84.tif"), 
            overwrite = TRUE)
writeRaster(ppt_wgs84, 
            filename = file.path(output_dir, "WGS84/ppt1991_2020_ave_CA_wgs84.tif"), 
            overwrite = TRUE)
writeRaster(cwd_wgs84, 
            filename = file.path(output_dir, "WGS84/cwd1991_2020_ave_CA_wgs84.tif"), 
            overwrite = TRUE)

# export gSSURGO rasters in WGS84
writeRaster(clay_wgs84, 
            filename = file.path(output_dir, "WGS84/clay_pct_0_10cm_CA_wgs84.tif"), 
            overwrite = TRUE)
writeRaster(ph_wgs84, 
            filename = file.path(output_dir, "WGS84/ph_0_10cm_CA_wgs84.tif"), 
            overwrite = TRUE)
writeRaster(ec_wgs84, 
            filename = file.path(output_dir, "WGS84/ec_dS_m_0_10cm_CA_wgs84.tif"), 
            overwrite = TRUE)

# export slope raster in WGS84
writeRaster(slope_wgs84, 
            filename = file.path(output_dir, "WGS84/slope_deg_CA_wgs84.tif"), 
            overwrite = TRUE)
```










