---
title: "Processing Env Predictor Variables for SDMs"
format: html
---

```{r}
# clear the environment
rm(list = ls())
```

## Overview/Purpose

This script processes environmental predictor variables for use in species distribution models (SDMs). It will load each of the predictor variables in raster format, ensure they are aligned to a common extent and resolution, and then export the processed rasters for use in modeling.

## Setup

#### Load packages

```{r}
# load required packages
library(terra)
library(here)
library(tidyverse)
library(tigris)
library(sf)
```

#### Load data

```{r}
# ==============================================================================
# BCM data 
# ==============================================================================

## AET
# file paths for BCM AET monthly average rasters
aet_path <- list.files(path = here("data/raw/bcm/bcmv8_historic/monthly_avgs/"), 
                       pattern = "aet_.*_avg\\.tif$", 
                       full.names = TRUE)

# load AET rasters as a SpatRaster
aet_stack <- rast(aet_path)


# ==============================================================================
# gSSURGO
# ==============================================================================

## % clay
# file path for gSSURGO % clay raster
clay_path <- here("data/raw/gssurgo/processed/clay_pct_0_10cm_sjv_270m.tif")
clay_rast <- rast(clay_path)

## pH
# file path for gSSURGO pH raster
ph_path <- here("data/raw/gssurgo/processed/pH_0_10cm_sjv_270m.tif")
ph_rast <- rast(ph_path)

# EC
# file path for gSSURGO EC raster
ec_path <- here("data/raw/gssurgo/processed/ec_dS_m_0_10cm_sjv_270m.tif")
ec_rast <- rast(ec_path)


# ==============================================================================
# slope (open topography)
# ==============================================================================

elevation_path <- here("data/raw/open_topography/output_USGS30m.tif")
elevation_rast <- rast(elevation_path)
```

## Data Processing

#### BCM

```{r}
# ==============================================================================
# AET
# ==============================================================================

# check the projection, extent, and resolution of the AET stack
crs(aet_stack) # EPSG:3310
ext(aet_stack) 
res(aet_stack) # 270m

# this is already in the desired projection, extent, and resolution
# we will resample/project other layers to match this one

# calculate the mean AET across all months
aet_mean <- mean(aet_stack)
plot(aet_mean, main = "Mean AET")
```

#### Make SJV spat vector

```{r}
# get California counties from Census TIGER/Line shapefiles
ca_counties <- counties(state = "CA")

# define San Joaquin Valley counties
sjv_counties <- c("San Joaquin", "Stanislaus", "Merced", "Madera", 
                  "Fresno", "Kings", "Tulare", "Kern")

# filter and combine into one polygon
sjv <- ca_counties %>%
  filter(NAME %in% sjv_counties) %>%
  st_union() %>% 
  st_transform(crs = crs(aet_mean)) %>% # transform to aet raster CRS
  vect() # convert to terra vector format

# crop AET mean raster to SJV extent
aet_mean_sjv <- aet_mean %>% 
  crop(sjv) %>%
  mask(sjv)

# plot cropped AET raster
plot(aet_mean_sjv, main = "Mean AET (SJV)")
```

#### Fill NA values in AET raster

The AET raster has some NA values in small areas. This causes errors in `Wallace`. We will fill these values with the mean value of neighboring cells.

```{r}
# define function to fill NA values with mean of neighboring cells
fill_all_na <- function(raster, max_iter = 100) {
  r <- raster
  for (i in 1:max_iter) {
    if (!any(is.na(values(r)))) break
    r <- focal(r, 
               w = 3,                # 3x3 moving window
               fun = mean, 
               na.rm = TRUE,         # ignore NAs in calculation
               na.policy = "only")   # only replace NA cells, leave others as is
  }
  return(r)
}

# apply function to AET raster
aet_mean_sjv_filled <- fill_all_na(aet_mean_sjv)

# crop to SJV again to ensure no edge effects
aet_mean_sjv_filled <- aet_mean_sjv_filled %>%
  crop(sjv) %>%
  mask(sjv)

# plot final AET raster
plot(aet_mean_sjv_filled, main = "Mean AET (filled NA values)")
```

#### gSSURGO

Resample gSSURGO rasters to match AET `ext`, `res`, and `crs`.

```{r}
# resample gSSURGO rasters to match AET
clay_resampled <- resample(clay_rast, 
                           aet_mean_sjv, 
                           method = "mean")
ph_resampled <- resample(ph_rast, 
                         aet_mean_sjv, 
                         method = "mean")
ec_resampled <- resample(ec_rast, 
                         aet_mean_sjv, 
                         method = "mean")

# confirm all rasters have the same extent and resolution
crs(clay_resampled) == crs(aet_mean_sjv)
crs(ph_resampled) == crs(aet_mean_sjv)
crs(ec_resampled) == crs(aet_mean_sjv)

ext(clay_resampled) == ext(aet_mean_sjv)
ext(ph_resampled) == ext(aet_mean_sjv)
ext(ec_resampled) == ext(aet_mean_sjv)

res(clay_resampled) == res(aet_mean_sjv)
res(ph_resampled) == res(aet_mean_sjv)
res(ec_resampled) == res(aet_mean_sjv)
```

The gSSURGO data has a lot of NA values in some areas. This causes errors in `Wallace`. We might fill these values with other data later, but for now we will fill them with the mean value of neighboring cells.

```{r}
# apply fill NA function to gSSURGO rasters
clay_filled <- fill_all_na(clay_resampled)
ph_filled <- fill_all_na(ph_resampled)
ec_filled <- fill_all_na(ec_resampled)

# crop to SJV again to ensure no edge effects
clay_filled <- clay_filled %>% 
  crop(sjv) %>%
  mask(sjv)
ph_filled <- ph_filled %>% 
  crop(sjv) %>%
  mask(sjv)
ec_filled <- ec_filled %>% 
  crop(sjv) %>%
  mask(sjv)

# make sure all rasters have the same extent and resolution
crs(clay_filled) == crs(aet_mean_sjv)
crs(ph_filled) == crs(aet_mean_sjv)
crs(ec_filled) == crs(aet_mean_sjv)

ext(clay_filled) == ext(aet_mean_sjv)
ext(ph_filled) == ext(aet_mean_sjv)
ext(ec_filled) == ext(aet_mean_sjv)

res(clay_filled) == res(aet_mean_sjv)
res(ph_filled) == res(aet_mean_sjv)
res(ec_filled) == res(aet_mean_sjv)

# plot filled gSSURGO rasters
plot(clay_filled, main = "% Clay (filled NA values)")
plot(ph_filled, main = "pH (filled NA values)")
plot(ec_filled, main = "EC (filled NA values)")
```


#### Slope

```{r}
# check the projection, extent, and resolution of the elevation raster
crs(elevation_rast) # EPSG:4326
ext(elevation_rast)
res(elevation_rast) # 30m


# project elevation raster to match AET and crop to SJV extent 
elevation_proj <- project(elevation_rast, 
                          aet_mean_sjv, 
                          method = "mean")
crs(elevation_proj) # confirm EPSG:3310


# calculate slope from elevation
slope_rast <- terrain(elevation_proj, 
                      v = "slope", 
                      unit = "degrees")
plot(slope_rast, main = "Slope (degrees)")


# crop to SJV
slope_sjv <- slope_rast %>% 
  crop(sjv) %>%
  mask(sjv)


# resample to match AET resolution and extent
slope_resampled <- resample(slope_sjv, 
                            aet_mean_sjv, 
                            method = "mean")
plot(slope_resampled, main = "Elevation")


# confirm all rasters have the same extent and resolution
crs(slope_resampled) == crs(aet_mean_sjv)
ext(slope_resampled) == ext(aet_mean_sjv)
res(slope_resampled) == res(aet_mean_sjv)
```

## Export Processed Rasters

#### EPSG:3310, 270m resolution, SJV extent

```{r}
# path to output directory
output_dir <- here("data/intermediate/misc/habitat_suitability/env_predictors/")

# export mean AET raster
writeRaster(aet_mean_sjv_filled, 
            filename = file.path(output_dir, "aet_mon_mean_sjv_270m.tif"), 
            overwrite = TRUE)

# export gSSURGO rasters
writeRaster(clay_filled, 
            filename = file.path(output_dir, "clay_pct_0_10cm_sjv_270m.tif"), 
            overwrite = TRUE)
writeRaster(ph_filled, 
            filename = file.path(output_dir, "ph_0_10cm_sjv_270m.tif"), 
            overwrite = TRUE)
writeRaster(ec_filled, 
            filename = file.path(output_dir, "ec_dS_m_0_10cm_sjv_270m.tif"), 
            overwrite = TRUE)

# export slope raster
writeRaster(slope_resampled, 
            filename = file.path(output_dir, "slope_deg_sjv_270m.tif"), 
            overwrite = TRUE)
```

#### WGS84, ~0.00242 degree resolution at equator, SJV extent

Reproject all rasters to WGS 84 (EPSG:4326) for use in Wallace.

```{r}
# project all rasters to WGS 84
aet_wgs84 <- project(aet_mean_sjv_filled, 
                     "EPSG:4326", 
                     method = "mean")
clay_wgs84 <- project(clay_filled, 
                      "EPSG:4326", 
                      method = "mean")
ph_wgs84 <- project(ph_filled, 
                    "EPSG:4326", 
                    method = "mean")
ec_wgs84 <- project(ec_filled, 
                    "EPSG:4326", 
                    method = "mean")
slope_wgs84 <- project(slope_resampled, 
                       "EPSG:4326", 
                       method = "mean")

# confirm all rasters have the same extent and resolution
crs(aet_wgs84) == crs(clay_wgs84)
crs(aet_wgs84) == crs(ph_wgs84)
crs(aet_wgs84) == crs(ec_wgs84)
crs(aet_wgs84) == crs(slope_wgs84)
ext(aet_wgs84) == ext(clay_wgs84)
ext(aet_wgs84) == ext(ph_wgs84)
ext(aet_wgs84) == ext(ec_wgs84)
ext(aet_wgs84) == ext(slope_wgs84)
res(aet_wgs84) == res(clay_wgs84)
res(aet_wgs84) == res(ph_wgs84)
res(aet_wgs84) == res(ec_wgs84)
res(aet_wgs84) == res(slope_wgs84)
```

Save the WGS84 rasters.

```{r}
# export mean AET raster in WGS84
writeRaster(aet_wgs84, 
            filename = file.path(output_dir, "WGS84/aet_mon_mean_sjv_wgs84.tif"), 
            overwrite = TRUE)

# export gSSURGO rasters in WGS84
writeRaster(clay_wgs84, 
            filename = file.path(output_dir, "WGS84/clay_pct_0_10cm_sjv_wgs84.tif"), 
            overwrite = TRUE)
writeRaster(ph_wgs84, 
            filename = file.path(output_dir, "WGS84/ph_0_10cm_sjv_wgs84.tif"), 
            overwrite = TRUE)
writeRaster(ec_wgs84, 
            filename = file.path(output_dir, "WGS84/ec_dS_m_0_10cm_sjv_wgs84.tif"), 
            overwrite = TRUE)

# export slope raster in WGS84
writeRaster(slope_wgs84, 
            filename = file.path(output_dir, "WGS84/slope_deg_sjv_wgs84.tif"), 
            overwrite = TRUE)
```











