---
title: "2_env_pred_processing_future"
format: html
---

```{r}
# clear the environment
rm(list = ls())
```

## Overview/Purpose

This script processes all of the future climate variables from the BCM that we'll use as environmental predictor variables to transfer our species distribution models (SDMs). We'll be processing the following data:

- `aet`: actual evapotranspiration
- `cwd`: climatic water deficit
- `ppt`: precipitation
- `tmn`: minimum temperature
- `tmx`: maximum temperature

For each climate variable, we have data for:

- two 30-year averaged time periods (`2020-2049`, `2040-2069`) 
- four GCMs (`CanESM2`, `CNRM-CM5`, `HadGEM2-ES`, `MIROC5`)
- two emission scenarios (`RCP45`, `RCP85`)

The resulting processed raster files will be saved in the `data/intermediate/misc/env_predictors/bcm_gcm_ensemble/` directory. There will be 12 files produced, one for each combination of climate variable, time period, and emission scenario averaged across all three GCMs.

## Setup

#### Load packages

```{r}
# load required packages
library(terra)
library(here)
library(tidyverse)
library(tigris)
library(sf)
library(tmap)
```

#### Load the template raster to match crs, extent, and resolution

```{r}
# processed aet historic raster
ref_raster <- rast(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline/aet1991_2020_ave_CA_270m.tif"))
```

#### Define future BCM data file paths & parameters

```{r}
# BCM future data path
input_dir <- here("data/raw/bcm/bcmv8_future/")
output_dir <- here("data/intermediate/misc/habitat_suitability/env_predictors/future/bcm_gcm_ensemble/")

# define parameters
climate_vars <- c("aet", "cwd", "ppt", "tmn", "tmx")
gcms <- c("CanESM2", "CNRM-CM5", "HadGEM2-ES", "MIROC5")
scenarios <- c("RCP45", "RCP85")
time_periods <- c("2020_2049", "2040_2069")
```

## Create functions for data processing

#### Function to create file path

```{r}
# file path function to get the .asc file path for a given: 
# variable, gcm, scenario, and period

get_asc_path <- function(input_dir, var, gcm, scenario, period) {
  # construct folder name: var_GCM_scenario_30yr
  folder_name <- paste0(var, "_", gcm, "_", scenario, "_30yr")
  
  # construct file name: var_period_ave.asc
  file_name <- paste0(var, "_", period, "_ave.asc")
  
  # full path
  file_path <- file.path(input_dir, folder_name, file_name)
  
  return(file_path)
}
```

#### Function to average rasters across GCMs

```{r}
# average rasters across GCMs for a given variable, scenario, and period

average_across_gcms <- function(input_dir, var, gcms, scenario, period) {
  # collect rasters from all GCMs
  raster_list <- list()
  
  for (i in seq_along(gcms)) {
    gcm <- gcms[i]
    file_path <- get_asc_path(input_dir, var, gcm, scenario, period)
    
    # check if file exists
    if (!file.exists(file_path)) {
      warning("File not found: ", file_path)
      next
    }
    
    # read the raster
    message("  Reading: ", basename(file_path), " from ", gcm)
    r <- rast(file_path)
    
    # set CRS - BCM data uses California Albers (EPSG:3310)
    crs(r) <- "EPSG:3310"
    
    # replace -9999 with NA
    message("    Converting -9999 to NA...")
    r[r == -9999] <- NA
    
    raster_list[[i]] <- r
  }
  
  # check if we have any rasters to average
  if (length(raster_list) == 0) {
    stop("No rasters found for ", var, " ", scenario, " ", period)
  }
  
  # stack all rasters and calculate mean
  raster_stack <- rast(raster_list)
  ensemble_mean <- mean(raster_stack, na.rm = TRUE)
  
  return(ensemble_mean)
}
```

#### Function to fill NAs with 3x3 focal mean

```{r}
# function to fill NA values with mean of neighboring cells (3x3 moving window)
fill_all_na <- function(raster, max_iter = 100) {
  r <- raster
  for (i in 1:max_iter) {
    if (!any(is.na(values(r)))) break
    r <- focal(r,
               w = 3,                # 3x3 moving window
               fun = mean,
               na.rm = TRUE,         # ignore NAs in calculation
               na.policy = "only")   # only replace NA cells, leave others as is
  }
  return(r)
}
```

#### Function to align raster to match processed historic raster

```{r}
# reproject and resample to match reference raster

align_to_reference <- function(input_raster, ref_raster, method = "mean") {
 
  # step 1: reproject to match reference CRS if needed
  if (!same.crs(input_raster, ref_raster)) {
    message("    Reprojecting to match reference CRS...")
    input_raster <- project(input_raster, crs(ref_raster))
  }
  
  # step 2: resample to match reference resolution and extent
  # this also effectively clips to the reference extent
  message("    Resampling to match reference resolution and extent...")
  aligned_raster <- resample(input_raster, ref_raster, method = method)
  
  # Step 3: sask to reference raster (use NA values from reference as mask)
  message("    Masking to reference raster...")
  aligned_raster <- mask(aligned_raster, ref_raster)
  
  return(aligned_raster)
}
```

## Process data: loop through variables, scenarios, and periods

```{r}
# track output files
output_files <- character()

# process each combination of variable, scenario, and time period
for (var in climate_vars) {
  for (scenario in scenarios) {
    for (period in time_periods) {
      
      message("\nProcessing: ", toupper(var), " | ", scenario, " | ", period)
      message(paste(rep("-", 50), collapse = ""))
      
      # calculate ensemble mean across GCMs
      ensemble_raster <- average_across_gcms(
        input_dir = input_dir,
        var = var,
        gcms = gcms,
        scenario = scenario,
        period = period
      )
      
      # align to reference raster (reproject, resample, clip, mask)
      message("  Aligning to reference raster...")
      ensemble_raster <- align_to_reference(
        input_raster = ensemble_raster,
        ref_raster = ref_raster,
        method = "mean" 
      )
      
      
      # fill remaining NAs with 3x3 moving window mean
      message("  Filling NAs with 3x3 moving window mean...")
      na_count_before <- sum(is.na(values(ensemble_raster)))
      message("    NAs before filling: ", na_count_before)
      
      ensemble_raster <- fill_all_na(ensemble_raster, max_iter = 100)
      
      na_count_after <- sum(is.na(values(ensemble_raster)))
      message("    NAs after filling: ", na_count_after)
      message("    NAs filled: ", na_count_before - na_count_after)
      
      
      # crop to ref raster to remove edge effects
      message("  Cropping to reference raster extent to remove edge effects...")
      ensemble_raster <- ensemble_raster %>% 
        crop(ref_raster) %>% 
        mask(ref_raster)
      
      
      # re-align again to reference
      message("  Re-aligning to reference raster after filling NAs...")
      ensemble_raster <- align_to_reference(
        input_raster = ensemble_raster,
        ref_raster = ref_raster,
        method = "mean"
      )
      
      # check final NA count
      final_na_count <- sum(is.na(values(ensemble_raster)))
      message("    Final NAs after re-alignment: ", final_na_count)
      
      
      # construct output filename
      # format: var_scenario_period_gcm_ensemble.tif
      output_name <- paste0(var, "_", scenario, "_", period, "_gcm_ensemble.tif")
      output_path <- file.path(output_dir, output_name)
      
      # save as GeoTIFF
      writeRaster(
        ensemble_raster,
        filename = output_path,
        overwrite = TRUE
      )
      
      message("  Saved: ", output_name)
      
      # get summary statistics
      message("  Summary statistics:")
      raster_summary <- summary(values(ensemble_raster))
      print(raster_summary)
      message("")
      
      output_files <- c(output_files, output_path)
    }
  }
}
```

#### Project to WGS84 (`EPSG:4326`) for use in Wallace

```{r}
# # define output directory for WGS84 files
# wgs_output_dir <- file.path(output_dir, "WGS84")
# 
# # track WGS84 output files
# wgs_output_files <- character()
# 
# # reproject each ensemble file to WGS84
# for (f in output_files) {
#   message("Reprojecting: ", basename(f))
#   
#   # load the raster
#   r <- rast(f)
#   
#   # reproject to WGS84
#   r_wgs <- project(r, "EPSG:4326", method = "mean")
#   
#   # construct output filename (add _wgs84 suffix)
#   original_name <- tools::file_path_sans_ext(basename(f))
#   wgs_output_name <- paste0(original_name, "_wgs84.tif")
#   wgs_output_path <- file.path(wgs_output_dir, wgs_output_name)
#   
#   # save as GeoTIFF
#   writeRaster(
#     r_wgs,
#     filename = wgs_output_path,
#     overwrite = TRUE
#   )
#   
#   message("  Saved: ", wgs_output_name)
#   wgs_output_files <- c(wgs_output_files, wgs_output_path)
# }
```

## Plot some of the projections

#### Load the SJV & CA boundaries for plotting

```{r}
# load SJV boundary for masking
sjv_boundary <- read_sf(here("data/raw/sjv_counties/sjv_counties.shp"))

# transform SJV boundary to EPSG:3310
sjv_boundary <- st_transform(sjv_boundary, crs = crs(ref_raster))
crs(sjv_boundary)
# convert to terra vector
sjv_vect <- vect(sjv_boundary)


# load CA boundary for inset map
ca_boundary <- read_sf(here("data/raw/ca_state/ca_state_boundary_wgs84.shp"))

# transform CA boundary to EPSG:3310
ca_boundary <- st_transform(ca_boundary, crs = crs(ref_raster))
crs(ca_boundary)
# convert to terra vector
ca_vect <- vect(ca_boundary)
```

#### Load the projections as raster stacks & crop to SJV

```{r}
# RCP45 (2020-2049) ensemble mean for each variable 
rcp45_2020_2049_path <- list.files(output_dir, 
                                   pattern = "RCP45_2020_2049.*\\.tif$",
                                   full.names = TRUE)
rcp45_2020_2049_stack <- rast(rcp45_2020_2049_path)

# RCP45 (2040-2069) ensemble mean for each variable
rcp45_2040_2069_path <- list.files(output_dir,
                                   pattern = "RCP45_2040_2069.*\\.tif$",
                                   full.names = TRUE)
rcp45_2040_2069_stack <- rast(rcp45_2040_2069_path)

# RCP85 (2020-2049) ensemble mean for each variable
rcp85_2020_2049_path <- list.files(output_dir,
                                   pattern = "RCP85_2020_2049.*\\.tif$",
                                   full.names = TRUE)
rcp85_2020_2049_stack <- rast(rcp85_2020_2049_path)

# RCP85 (2040-2069) ensemble mean for each variable
rcp85_2040_2069_path <- list.files(output_dir,
                                   pattern = "RCP85_2040_2069.*\\.tif$",
                                   full.names = TRUE)
rcp85_2040_2069_stack <- rast(rcp85_2040_2069_path)


# crop to SJV extent
rcp45_2020_2049_stack_sjv <- rcp45_2020_2049_stack %>% 
  crop(sjv_vect) %>% 
  mask(sjv_vect)
rcp45_2040_2069_stack_sjv <- rcp45_2040_2069_stack %>% 
  crop(sjv_vect) %>% 
  mask(sjv_vect)
rcp85_2020_2049_stack_sjv <- rcp85_2020_2049_stack %>% 
  crop(sjv_vect) %>% 
  mask(sjv_vect)
rcp85_2040_2069_stack_sjv <- rcp85_2040_2069_stack %>% 
  crop(sjv_vect) %>% 
  mask(sjv_vect)
```

#### Plot the projections

```{r}
# convert terra vectors to sf for easier handling
sjv_sf <- st_as_sf(sjv_vect)
ca_sf <- st_as_sf(ca_vect)

## Create expanded bounding box for SJV
sjv_bbox <- st_bbox(sjv_sf)
sjv_buffer <- 0.1
x_expand <- sjv_buffer * (unname(sjv_bbox["xmax"]) - unname(sjv_bbox["xmin"]))
y_expand <- sjv_buffer * (unname(sjv_bbox["ymax"]) - unname(sjv_bbox["ymin"]))

sjv_bbox_expanded <- st_bbox(c(
  xmin = unname(sjv_bbox["xmin"]) - x_expand,
  xmax = unname(sjv_bbox["xmax"]) + x_expand,
  ymin = unname(sjv_bbox["ymin"]) - y_expand,
  ymax = unname(sjv_bbox["ymax"]) + y_expand
), crs = st_crs(sjv_sf))

# Create expanded bounding box for CA
ca_bbox <- st_bbox(ca_sf)
ca_buffer <- 0.1
x_expand_ca <- ca_buffer * (unname(ca_bbox["xmax"]) - unname(ca_bbox["xmin"]))
y_expand_ca <- ca_buffer * (unname(ca_bbox["ymax"]) - unname(ca_bbox["ymin"]))

ca_bbox_expanded <- st_bbox(c(
  xmin = unname(ca_bbox["xmin"]) - x_expand_ca,
  xmax = unname(ca_bbox["xmax"]) + x_expand_ca,
  ymin = unname(ca_bbox["ymin"]) - y_expand_ca,
  ymax = unname(ca_bbox["ymax"]) + y_expand_ca
), crs = st_crs(ca_sf))
```

```{r}
# stack the RCP45 and RCP85 stacks together for plotting and name them
rcp45_stack <- c(rcp45_2020_2049_stack_sjv, rcp45_2040_2069_stack_sjv)
names(rcp45_stack) <- c("AET RCP45 (2020_2049)", "CWD RCP45 (2020_2049)",
                        "PPT RCP45 (2020_2049)", "TMN RCP45 (2020_2049)", 
                        "TMX RCP45 (2020_2049)",
                        "AET RCP45 (2040_2069)", "CWD RCP45 (2040_2069)", 
                        "PPT RCP45 (2040_2069)", "TMN RCP45 (2040_2069)", 
                        "TMX RCP45 (2040_2069)")

rcp85_stack <- c(rcp85_2020_2049_stack_sjv, rcp85_2040_2069_stack_sjv)
names(rcp85_stack) <- c("AET RCP85 (2020_2049)", "CWD RCP85 (2020_2049)",
                        "PPT RCP85 (2020_2049)", "TMN RCP85 (2020_2049)", 
                        "TMX RCP85 (2020_2049)",
                        "AET RCP85 (2040_2069)", "CWD RCP85 (2040_2069)", 
                        "PPT RCP85 (2040_2069)", "TMN RCP85 (2040_2069)",
                        "TMX RCP85 (2040_2069)")

# plot the stacks using tmap
tm_shape(rcp45_stack, bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"), 
            col.legend = tm_legend(title = "mm / yr", 
                                   position = tm_pos_in("left", "bottom"))) +
  tm_layout(panel.labels = names(rcp45_stack), 
            panel.label.size = 0.8) +
  tm_facets(ncol = 3)
```

## Raster math between the future projections

#### Find the difference between RCP45 projections

Interpretation:
- Positive values indicate an increase in `mm/yr` from the earlier period (2020-2049) to the later period (2040-2069).
- Negative values indicate a decrease in `mm/yr` from the earlier period to the later period.

```{r}
# calculate difference between RCP45 projections for each variable and time period
aet_diff_rcp45_2020_2049 <- rcp45_stack[[4]] - rcp45_stack[[1]]
cwd_diff_rcp45_2020_2049 <- rcp45_stack[[5]] - rcp45_stack[[2]]
ppt_diff_rcp45_2020_2049 <- rcp45_stack[[6]] - rcp45_stack[[3]]

# plot the differences
rcp45_diff_stack <- c(aet_diff_rcp45_2020_2049, cwd_diff_rcp45_2020_2049,
                      ppt_diff_rcp45_2020_2049)
names(rcp45_diff_stack) <- c("AET Δ (2040-2069 - 2020-2049)",
                             "CWD Δ (2040-2069 - 2020-2049)",
                             "PPT Δ (2040-2069 - 2020-2049)")

tm_shape(rcp45_diff_stack, bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "RdBu"),
            col.legend = tm_legend(title = "mm / yr",
                                   position = tm_pos_in("left", "bottom"))) +
  tm_layout(panel.labels = names(rcp45_diff_stack),
            panel.label.size = 1) +
  tm_facets(ncol = 3)
```

#### Find the difference between RCP85 (2040-2069) & RCP45 (2020-2049) projections

```{r}
# calculate difference between RCP85 (2040-2069) and RCP45 (2020-2049) 
aet_diff_rcp85_2040_2069_vs_rcp45_2020_2049 <- rcp85_stack[[4]] - rcp45_stack[[1]]
cwd_diff_rcp85_2040_2069_vs_rcp45_2020_2049 <- rcp85_stack[[5]] - rcp45_stack[[2]]
ppt_diff_rcp85_2040_2069_vs_rcp45_2020_2049 <- rcp85_stack[[6]] - rcp45_stack[[3]]

# plot the differences
rcp85_vs_rcp45_diff_stack <- c(aet_diff_rcp85_2040_2069_vs_rcp45_2020_2049, 
                               cwd_diff_rcp85_2040_2069_vs_rcp45_2020_2049,
                               ppt_diff_rcp85_2040_2069_vs_rcp45_2020_2049)
names(rcp85_vs_rcp45_diff_stack) <- c("AET Δ (RCP85 2040-2069 - RCP45 2020-2049)", 
                                      "CWD Δ (RCP85 2040-2069 - RCP45 2020-2049)", 
                                      "PPT Δ (RCP85 2040-2069 - RCP45 2020-2049)")

tm_shape(rcp85_vs_rcp45_diff_stack, bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "RdBu"), 
            col.legend = tm_legend(title = "mm / yr", 
                                   position = tm_pos_in("left", "bottom"))) +
  tm_layout(panel.labels = names(rcp85_vs_rcp45_diff_stack), 
            panel.label.size = 0.8) +
  tm_facets(ncol = 3)
```

## Raster math between future and baseline projections

#### Find the difference between future and baseline BCM data

Load the processed baseline (1991-2020) BCM rasters

```{r}
# load the 1991-2020 average rasters in EPSG 3310
aet_1991_2020 <- rast(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline/aet1991_2020_ave_CA_270m.tif"))
cwd_1991_2020 <- rast(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline/cwd1991_2020_ave_CA_270m.tif"))
ppt_1991_2020 <- rast(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline/ppt1991_2020_ave_CA_270m.tif"))
tmn_1991_2020 <- rast(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline/tmn1991_2020_ave_CA_270m.tif"))
tmx_1991_2020 <- rast(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline/tmx1991_2020_ave_CA_270m.tif"))

# crop rasters to SJV boundary vect
aet_1991_2020_sjv <- aet_1991_2020 %>%
  crop(sjv_vect) %>%
  mask(sjv_vect)
cwd_1991_2020_sjv <- cwd_1991_2020 %>%
  crop(sjv_vect) %>%
  mask(sjv_vect)
ppt_1991_2020_sjv <- ppt_1991_2020 %>%
  crop(sjv_vect) %>%
  mask(sjv_vect)
tmn_1991_2020_sjv <- tmn_1991_2020 %>%
  crop(sjv_vect) %>%
  mask(sjv_vect)
tmx_1991_2020_sjv <- tmx_1991_2020 %>%
  crop(sjv_vect) %>%
  mask(sjv_vect)
```

Calculate the difference between RCP45 projections and baseline for each variable and time period.

```{r}
# calculate difference between RCP45 projections and baseline
aet_diff_rcp45_2020_2049_vs_baseline <- rcp45_stack[[1]] - aet_1991_2020_sjv
cwd_diff_rcp45_2020_2049_vs_baseline <- rcp45_stack[[2]] - cwd_1991_2020_sjv
ppt_diff_rcp45_2020_2049_vs_baseline <- rcp45_stack[[3]] - ppt_1991_2020_sjv
tmn_diff_rcp45_2020_2049_vs_baseline <- rcp45_stack[[4]] - tmn_1991_2020_sjv
tmx_diff_rcp45_2020_2049_vs_baseline <- rcp45_stack[[5]] - tmx_1991_2020_sjv

aet_diff_rcp45_2040_2069_vs_baseline <- rcp45_stack[[6]] - aet_1991_2020_sjv
cwd_diff_rcp45_2040_2069_vs_baseline <- rcp45_stack[[7]] - cwd_1991_2020_sjv
ppt_diff_rcp45_2040_2069_vs_baseline <- rcp45_stack[[8]] - ppt_1991_2020_sjv
tmn_diff_rcp45_2040_2069_vs_baseline <- rcp45_stack[[9]] - tmn_1991_2020_sjv
tmx_diff_rcp45_2040_2069_vs_baseline <- rcp45_stack[[10]] - tmx_1991_2020_sjv
```

Plot the raw baseline rasters

```{r}
# plot each raster and then combine into a multi-panel plot

# AET
aet_base_plot <- tm_shape(aet_1991_2020_sjv, bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"), 
            col.legend = tm_legend(title = "mm / yr", 
                                   position = tm_pos_in("left", "bottom"))) + 
  tm_title("AET (1991-2020 avg)", 
           position = tm_pos_in("right", "top"), 
           size = 0.8)

# CWD
cwd_base_plot <- tm_shape(cwd_1991_2020_sjv, bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
            col.legend = tm_legend(title = "mm / yr", 
                                   position = tm_pos_in("left", "bottom"))) + 
  tm_title("CWD (1991-2020 avg)", 
           position = tm_pos_in("right", "top"), 
           size = 0.8)

# PPT
ppt_base_plot <- tm_shape(ppt_1991_2020_sjv, bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
            col.legend = tm_legend(title = "mm / yr", 
                                   position = tm_pos_in("left", "bottom"))) +
  tm_title("PPT (1991-2020 avg)", 
           position = tm_pos_in("right", "top"), 
           size = 0.8)

# TMN
tmn_base_plot <- tm_shape(tmn_1991_2020_sjv, bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
            col.legend = tm_legend(title = "°C", 
                                   position = tm_pos_in("left", "bottom"))) +
  tm_title("TMN (1991-2020 avg)", 
           position = tm_pos_in("right", "top"), 
           size = 0.8)

# TMX
tmx_base_plot <- tm_shape(tmx_1991_2020_sjv, bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
            col.legend = tm_legend(title = "°C", 
                                   position = tm_pos_in("left", "bottom"))) +
  tm_title("TMX (1991-2020 avg)", 
           position = tm_pos_in("right", "top"), 
           size = 0.8)

# combine into multi-panel plot
base_plot <- tmap_arrange(aet_base_plot, cwd_base_plot, ppt_base_plot, 
                              tmn_base_plot, tmx_base_plot, 
                              ncol = 3)

# create the inset map
inset_map <- tm_shape(ca_vect, bbox = ca_bbox_expanded) +
  tm_borders(col = "black", lwd=1) +
  tm_shape(sjv_vect) +
  tm_fill(fill = "red", fill_alpha = 0.5) +
  tm_borders(col = "red") +
  tm_layout(frame = TRUE, bg.color = "white")

# print main map
print(base_plot)

# add inset using viewport
print(inset_map, vp = grid::viewport(x = 0.75, y = 0.36, 
                                     width = 0.2, height = 0.25))
```

Plot the differences between RCP45 (2020-2049) projections and baseline

Interpretation (`future - baseline`):

- **Positive values** indicate an increase in the variable (e.g., more AET, more PPT, higher temperatures) in the future period compared to the baseline.
- **Negative values** indicate a decrease in the variable (e.g., less AET, less PPT, lower temperatures) in the future period compared to the baseline.

```{r}
# plot each difference raster and then combine into a multi-panel plot

# AET
aet_diff_plot <- tm_shape(aet_diff_rcp45_2020_2049_vs_baseline, 
                          bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "RdBu"), 
            col.legend = tm_legend(title = "mm / yr", 
                                   position = tm_pos_in("left", "bottom"))) + 
  tm_title("AET Δ (RCP45 2020-2049 - baseline)", 
           position = tm_pos_in("right", "top"), 
           size = 1)

# CWD
cwd_diff_plot <- tm_shape(cwd_diff_rcp45_2020_2049_vs_baseline, 
                          bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "RdBu"),
            col.legend = tm_legend(title = "mm / yr", 
                                   position = tm_pos_in("left", "bottom"))) + 
  tm_title("CWD Δ (RCP45 2020-2049 - baseline)", 
           position = tm_pos_in("right", "top"), 
           size = 1)

# PPT
ppt_diff_plot <- tm_shape(ppt_diff_rcp45_2020_2049_vs_baseline, 
                          bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "RdBu"),
            col.legend = tm_legend(title = "mm / yr", 
                                   position = tm_pos_in("left", "bottom"))) +
  tm_title("PPT Δ (RCP45 2020-2049 - baseline)", 
           position = tm_pos_in("right", "top"), 
           size = 1)

# TMN
tmn_diff_plot <- tm_shape(tmn_diff_rcp45_2020_2049_vs_baseline, 
                          bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "RdBu"),
            col.legend = tm_legend(title = "°C", 
                                   position = tm_pos_in("left", "bottom"))) +
  tm_title("TMN Δ (RCP45 2020-2049 - baseline)", 
           position = tm_pos_in("right", "top"), 
           size = 1)

# TMX
tmx_diff_plot <- tm_shape(tmx_diff_rcp45_2020_2049_vs_baseline, 
                          bbox = sjv_bbox_expanded) +
  tm_raster(col.scale = tm_scale_continuous(values = "RdBu"),
            col.legend = tm_legend(title = "°C", 
                                   position = tm_pos_in("left", "bottom"))) +
  tm_title("TMX Δ (RCP45 2020-2049 - baseline)", 
           position = tm_pos_in("right", "top"), 
           size = 1)

# combine into multi-panel plot
diff_plot <- tmap_arrange(aet_diff_plot, cwd_diff_plot, ppt_diff_plot, 
                              tmn_diff_plot, tmx_diff_plot, 
                              ncol = 3)

# print main map
print(diff_plot)

# add inset using viewport
print(inset_map, vp = grid::viewport(x = 0.75, y = 0.36, 
                                     width = 0.2, height = 0.25))
```




