---
title: "Extract environmental data to BNLL occurrences"
author: "Maxwell Pepperdine"
format: html
editor: visual
date: last-modified
editor_options: 
  chunk_output_type: console
---

## Overview

This script extracts environmental data to BNLL occurrence points for use in MaxEnt modeling. The environmental data includes:

-   BCM climate (AET, PPT)
-   gSSURGO soil data (pH, % clay, EC)
-   Slope

We will also spatially thin the occurrence points by 1 km to reduce spatial autocorrelation.

```{r}
rm(list=ls())
```

## Load packages

```{r}
library(terra)      # for raster processing
library(tidyverse)  # always
library(sf)         # for spatial data processing
library(here)       # for file paths
library(spThin)     # for spatial thinning
```

## Load data

#### Environmental rasters

```{r}
# load all env rasters
env_rasters <- rast(list.files(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline/keep_NAs"), 
                               pattern = ".tif$", 
                               full.names = TRUE))
plot(env_rasters)
```

#### BNLL occurrences

```{r}
# load occurrence points
bnll_occ <- read_csv("data/raw/occurrence_points/bnll/wallace/bnll_occur_wallace.csv")

# convert occ points to SpatVector
bnll_occ_vect <- vect(bnll_occ, 
                      geom = c("longitude", "latitude"), 
                      crs = "EPSG:4326")

# make sure crs matches env rasters
bnll_occ_vect <- project(bnll_occ_vect, crs(env_rasters))
```

#### Spatial thinning of occurrence points

Thin occurrence points to 1 km minimum distance. This reduces spatial autocorrelation and sampling bias.

```{r}
# convert back to lat/lon for spThin (it requires decimal degrees)
bnll_occ_latlon <- project(bnll_occ_vect, "EPSG:4326")
bnll_occ_df <- as.data.frame(bnll_occ_latlon, geom = "XY")

# run spatial thinning with 1 km distance
# thin.par = 1 means 1 km minimum distance between points
# reps = 100 means run 100 iterations to find best thinning solution
# locs.thinned.list.return = TRUE returns all replicates
# write.files = FALSE prevents writing files automatically
thinned_data <- spThin::thin(
  loc.data = bnll_occ_df,
  lat.col = "y",  # latitude column
  long.col = "x", # longitude column
  spec.col = "scientific_name", 
  thin.par = 1,   # thinning distance in km
  reps = 100,     # number of replications
  locs.thinned.list.return = TRUE,
  write.files = FALSE,
  write.log.file = FALSE
)

# select the thinned dataset with the most points
# (spThin returns a list of different thinning solutions)
max_points <- which.max(sapply(thinned_data, nrow))
bnll_thinned_df <- thinned_data[[max_points]]

# convert thinned points back to SpatVector and match original CRS
bnll_thinned_vect <- vect(bnll_thinned_df, 
                          geom = c("Longitude", "Latitude"), 
                          crs = "EPSG:4326")

# project to match env rasters
bnll_thinned_vect <- project(bnll_thinned_vect, crs(env_rasters))
crs(bnll_thinned_vect)  # check crs
```

## Extract environmental data to occurrence points

```{r}
# extract env data to occ points
bnll_occ_env <- terra::extract(env_rasters, bnll_thinned_vect, 
                               bind = TRUE)

# set as sf object
bnll_occ_env_sf <- st_as_sf(bnll_occ_env, 
                            coords = c("longitude", "latitude"))

# extract geometry columns as lat/lon
bnll_occ_env_df <- bnll_occ_env_sf %>% 
  mutate(longitude = st_coordinates(bnll_occ_env_sf)[,1],
         latitude = st_coordinates(bnll_occ_env_sf)[,2], 
         .before = "aet1991_2020_ave") %>%
  st_drop_geometry()
```

## Save the data

```{r}
# save final occurrence points with env data to CSV
write_csv(bnll_occ_env_df, here("data/intermediate/misc/habitat_suitability/sdm_files/bnll_sdm/1_extract_occ_env_data/bnll_occ_env_data.csv"))

# save SF object
write_sf(bnll_occ_env_sf, here("data/intermediate/misc/habitat_suitability/sdm_files/bnll_sdm/1_extract_occ_env_data/bnll_occ_env_data.shp"),
         append = FALSE)
```


















