---
title: "Extract environmental data to background points"
format: html
editor: visual
date: last-modified
editor_options: 
  chunk_output_type: console
---

## Overview

This script extracts environmental data to the BNLL background points (created in `2_generate_bg_points.qmd`) for use in MaxEnt modeling. The environmental data includes:

-   BCM climate (AET, PPT)
-   gSSURGO soil data (pH, % clay, EC)
-   Slope

```{r}
rm(list=ls())
```

## Load packages

```{r}
library(terra)      # for raster processing
library(tidyverse)  # always
library(sf)         # for spatial data processing
library(here)       # for file paths
```

## Load data

#### Environmental rasters

```{r}
# load all env rasters
env_rasters <- rast(list.files(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline"), 
                               pattern = ".tif$", 
                               full.names = TRUE))
# fix names
names(env_rasters) <- c("aet1991_2020_ave", "clay_pct_0_10cm", "cwd1991_2020_ave", 
                        "ec_dS_m_0_10cm", "elevation", "pH_0_10cm", "ppt1991_2020_ave", 
                        "slope")
plot(env_rasters)
```

#### BNLL background points (created in `2_generate_bg_points.qmd`)

```{r}
# load background points
bnll_bg <- read_csv(here("data/intermediate/misc/habitat_suitability/sdm_files/bnll_sdm/2_generate_bg_points/bnll_bg_points.csv"))

# convert bg points to SpatVector
bnll_bg_vect <- vect(bnll_bg, 
                     geom = c("longitude", "latitude"), 
                     crs = "EPSG:4326")

# make sure crs matches env rasters
bnll_bg_vect <- project(bnll_bg_vect, crs(env_rasters))
crs(bnll_bg_vect)
```

#### Spatial thinning of background points (1 km)

Thin background points by 1 km to reduce spatial autocorrelation.

```{r}
# convert back to lat/lon for spThin (it requires decimal degrees)
bnll_bg_latlon <- project(bnll_bg_vect, "EPSG:4326")
bnll_bg_df <- as.data.frame(bnll_bg_latlon, geom = "XY")
# add species dummy column for spThin
bnll_bg_df$species <- "background"

# run spatial thinning with 1 km distance
thinned_data <- spThin::thin(
  loc.data = bnll_bg_df,
  lat.col = "y",  # latitude column
  long.col = "x", # longitude column
  spec.col = "species", # dummy species column
  thin.par = 1,   # thinning distance in km
  reps = 10,     # number of replications
  locs.thinned.list.return = TRUE,
  write.files = FALSE,
  write.log.file = FALSE
)

# select the thinned dataset with the most points
max_points <- which.max(sapply(thinned_data, nrow))
bnll_bg_thinned_df <- thinned_data[[max_points]]

# convert thinned points back to SpatVector and match original CRS
bnll_bg_thinned_vect <- vect(bnll_bg_thinned_df, 
                             geom = c("Longitude", "Latitude"), 
                             crs = "EPSG:4326")

# project to match env rasters
bnll_bg_thinned_vect <- project(bnll_bg_thinned_vect, crs(env_rasters))
crs(bnll_bg_thinned_vect)  # check crs
```

## Extract environmental data to background points

```{r}
# extract env data to bg points
bnll_bg_env <- terra::extract(env_rasters, bnll_bg_thinned_vect, 
                              bind = TRUE)

# set as sf object
bnll_bg_env_sf <- st_as_sf(bnll_bg_env, 
                           coords = c("longitude", "latitude")) %>% 
  dplyr::select(all_of(names(env_rasters)))

# extract geometry columns as lon/lat
bnll_bg_env_df <- bnll_bg_env_sf %>% 
  mutate(longitude = st_coordinates(bnll_bg_env_sf)[,1],
         latitude = st_coordinates(bnll_bg_env_sf)[,2], 
         .before = 1) %>% 
  st_drop_geometry()
```

## Save the data

```{r}
# save final background points with env data to CSV
write_csv(bnll_bg_env_df, here("data/intermediate/misc/habitat_suitability/sdm_files/bnll_sdm/3_extract_bg_env_data/bnll_bg_env_data.csv"))

# save SF object
write_sf(bnll_bg_env_sf, here("data/intermediate/misc/habitat_suitability/sdm_files/bnll_sdm/3_extract_bg_env_data/bnll_bg_env_data.shp"),
         append = FALSE)
```























