---
title: "Extract environmental data to background points"
format: html
editor: visual
date: last-modified
editor_options: 
  chunk_output_type: console
---

## Overview

This script extracts environmental data to the BNLL background points (created in `2_generate_bg_points_bnll.qmd`) for use in MaxEnt modeling. The environmental data includes:

-   BCM climate (AET, CWD, PPT, TMN, TMX)
-   gSSURGO soil data (pH, % clay, % silt, % sand, EC)
-   Slope from USGS Open Topography

```{r}
rm(list=ls())
```

## Load packages

```{r}
library(terra)      # for raster processing
library(tidyverse)  # always
library(sf)         # for spatial data processing
library(here)       # for file paths
```

## Load data

#### Environmental rasters

```{r}
# load all env rasters
env_rasters <- rast(list.files(here("data/intermediate/misc/habitat_suitability/env_predictors/baseline"), 
                               pattern = ".tif$", 
                               full.names = TRUE))
# fix names
names(env_rasters) <- c("aet1991_2020_ave", "clay_pct_0_10cm", "cwd1991_2020_ave", 
                        "ec_dS_m_0_10cm", "pet1991_2020_ave", "pH_0_10cm", "ppt1991_2020_ave", 
                        "sand_pct_0_10cm", "silt_pct_0_10cm", "slope", 
                        "tmn1991_2020_ave", "tmx1991_2020_ave")
#plot(env_rasters)
```

#### BNLL background points (created in `2_generate_bg_points_bnll.qmd`)

```{r}
# load background points
bnll_bg <- read_csv(here("data/intermediate/misc/habitat_suitability/sdm_files/bnll_sdm/2_generate_bg_points_bnll/bnll_bg_points.csv"))

# convert bg points to SpatVector
bnll_bg_vect <- vect(bnll_bg, 
                     geom = c("longitude", "latitude"), 
                     crs = "EPSG:4326")

# make sure crs matches env rasters
bnll_bg_vect <- project(bnll_bg_vect, crs(env_rasters))
crs(bnll_bg_vect)
```

#### Spatial thinning of background points (270 m)

Thin background points to one per 270 m raster cell to match occurrence thinning and environmental predictor resolution.

```{r}
# Thin background points to one per raster cell (270 m)
# get the raster cell ID for each background point
cell_ids <- terra::cellFromXY(env_rasters[[1]], crds(bnll_bg_vect))

# shuffle so the kept point per cell is random
set.seed(123)
random_order <- sample(seq_len(nrow(bnll_bg_vect)))
bnll_bg_shuffled <- bnll_bg_vect[random_order, ]
cell_ids_shuffled <- cell_ids[random_order]

# keep first (random) point per cell
bnll_bg_thinned_vect <- bnll_bg_shuffled[!duplicated(cell_ids_shuffled), ]
```

## Extract environmental data to background points

```{r}
# extract env data to bg points
bnll_bg_env <- terra::extract(env_rasters, bnll_bg_thinned_vect, 
                              bind = TRUE)

# set as sf object
bnll_bg_env_sf <- st_as_sf(bnll_bg_env, 
                           coords = c("longitude", "latitude")) %>% 
  dplyr::select(all_of(names(env_rasters)))

# extract geometry columns as lon/lat
bnll_bg_env_df <- bnll_bg_env_sf %>% 
  mutate(longitude = st_coordinates(bnll_bg_env_sf)[,1],
         latitude = st_coordinates(bnll_bg_env_sf)[,2], 
         .before = 1) %>% 
  st_drop_geometry()
```

## Save the data

```{r}
# save final background points with env data to CSV
write_csv(bnll_bg_env_df, here("data/intermediate/misc/habitat_suitability/sdm_files/bnll_sdm/3_extract_bg_env_data_bnll/bnll_bg_env_data.csv"))
```
