---
title: "Generate GKR background points"
author: "Stella Wing"
format: html
editor: visual
date: last-modified
editor_options: 
  chunk_output_type: console
---

```{r}
# clean environment
rm(list = ls())
```

## Overview

This script generates the bg data for habitat suitability modeling for the giant kangaroo rat (*Dipodomys ingens*) in the San Joaquin Valley, CA. It will load our GKR occurrence data, generate a 50 km buffer around each occurrence point, and then find all geographically unique vertebrate (phylum Vertebrata) occurrence points within that buffer. These will be the background points for the GKR MaxEnt model.

## Load packages

```{r}
library(tidyverse)  # always
library(here)       # for file paths
library(rgbif)      # for GBIF data download
library(sf)         # for spatial data processing
```

## Load data

```{r}
# Load SJFK occurrence data
sjkf_occs <- read_csv(here("data/sjkf_occ_env_data.csv"))

# CA boundary
ca_boundary <- read_sf(here("data/ca_state/ca_state_boundary_no_islands.shp"))
```

## Data processing

#### Create 50 km buffer around SJKF occ points

```{r}
# convert SJKF occ data to sf object
sjkf_sf <- st_as_sf(sjkf_occs, 
                    coords = c("longitude", "latitude"), 
                    crs = 3310)

# create 50km km buffer around each occurrence point
sjkf_buffer <- st_buffer(sjkf_sf, dist = 50000)

# combine buffers into a single polygon
sjkf_buffer_union <- st_union(sjkf_buffer)

# clip to CA boundary to remove ocean areas
sjkf_buffer_union <- st_intersection(sjkf_buffer_union, 
                                    st_transform(ca_boundary, crs = 3310)) 

# transform buffer back to WGS84
sjkf_buffer_wgs84 <- st_transform(sjkf_buffer_union, crs = 4326)

# # save buffer to file for testing
# write_sf(gkr_buffer_wgs84, here("data/intermediate/misc/habitat_suitability/sdm_files/gkr_sdm/2_generate_bg_points_gkr/gkr_100km_buffer.shp"), 
#          append = FALSE)
```

```{r}
# set buffer to WKT format for rgbif::occ_download()
sjkf_buffer_wkt <- st_as_text(st_geometry(sjkf_buffer_wgs84)[[1]])
nchar(sjkf_buffer_wkt)  # check length of WKT string


st_geometry_type(sjkf_buffer_wgs84)
st_is_valid(sjkf_buffer_wgs84)


# buffer is too long for rgbif, so simplify it
buffer_simplify <- st_simplify(sjkf_buffer_wgs84, dTolerance = 1000)
buffer_simplify_wkt <- st_as_text(st_geometry(buffer_simplify)[[1]])
nchar(buffer_simplify_wkt)  # check length of WKT string
```

#### Obtain & process occ records using `rgbif::occ_download()`

```{r}
# set up GBIF credentials (one-time setup)
options(gbif_user = "stellawing")
options(gbif_pwd = "Lire54321$")
options(gbif_email = "stellawing@ucsb.edu")

# use rgbif::occ_download to acquire background points
download_key <- occ_download(
  pred_in("taxonKey", 44),  # Vertebrata
  pred("hasCoordinate", TRUE),
  pred("hasGeospatialIssue", FALSE),
  pred_gte("year", 1990),
  pred_lte("year", 2025),
  pred_in("basisOfRecord", c("HUMAN_OBSERVATION", "MACHINE_OBSERVATION", "OBSERVATION")),
  pred_lte("coordinateUncertaintyInMeters", 200),
  pred_within(buffer_simplify_wkt)  # Your buffer
)

# check download status
cat("Download key:", download_key, "\n")
cat("Check status at: https://www.gbif.org/occurrence/download/", download_key, "\n")

# check the download status programatically; this might take a while...
occ_download_wait(download_key, status_ping = 10)  # Check every 10 seconds

# download the data
download_file <- occ_download_get(download_key, path = ".")

# import the data
vertebrate_data <- occ_download_import(download_file)

# convert to sf
vertebrate_sf <- st_as_sf(vertebrate_data,
                          coords = c("decimalLongitude", "decimalLatitude"),
                          crs = 4326) %>% 
  # remove all SJKF occ points
  filter(species != "Vulpes macrotis") %>% 
  filter(taxonKey != 6164308)


# intersect occ points with the original precise buffer
# this keeps only points within 50 km of SJKF occs
vertebrate_in_buffer <- st_intersection(vertebrate_sf, sjkf_buffer_wgs84)

# remove duplicates
vertebrate_unique <- vertebrate_in_buffer %>%
  distinct(geometry, .keep_all = TRUE) %>% 
  # save relevant columns only
  dplyr::select(gbifID, institutionCode, year, basisOfRecord, scientificName, class, 
         classKey, species)

# extract coordinates for MaxEnt and env data extraction
background_coords <- st_coordinates(vertebrate_unique)
colnames(background_coords) <- c("longitude", "latitude")

# Create final dataframe
background_df <- cbind(
  as.data.frame(st_drop_geometry(vertebrate_unique)),
  background_coords
)
```

```{r}
# checking  bg pointy locations 

library(terra)

env_rasters <- rast(list.files(here("data/baseline/sjkf"), 
                               pattern = ".tif$", 
                               full.names = TRUE))
# fix names
names(env_rasters) <- c("aet1991_2020_ave", "clay_pct_0_10cm", "cwd1991_2020_ave", 
                        "ec_dS_m_0_10cm", "pH_0_10cm", "ppt1991_2020_ave", 
                        "sand_pct_0_10cm", "silt_pct_0_10cm", "slope", 
                        "tmn1991_2020_ave", "tmx1991_2020_ave")

# Select precipitation raster
r <- env_rasters[["ppt1991_2020_ave"]]

# Convert background_df to SpatVector
bg_vect <- vect(background_df,
                geom = c("longitude", "latitude"),
                crs = "EPSG:4326")

# Reproject to match raster CRS
bg_vect <- project(bg_vect, crs(r))

# Plot raster
plot(r,
     main = "Background Points on Mean Annual Precipitation")

# Add background points
points(bg_vect,
       pch = 16,
       col = "red",
       cex = 0.4)

```

Save the data

```{r}
# save SF to test
write_sf(vertebrate_unique, here("data/sjkf_bg_points.shp"),
         append = FALSE)


# save final background points to CSV
write_csv(background_df, here("data/sjkf_bg_points.csv"))
```
